---
title: "Assignment 2 Q3"
author: "Fenglin Chen"
date: "27/10/2021"
output: beamer_presentation
---

## Table of Contents

- Summary
- Math formula
- Math example
- General code
- Code example

## Summary

- $a\left(\mathcal{P} \right)$ is an attribute calculated over a population
- Influence is how much $a\left(\mathcal{P} \right)$ changes when one unit $u$ is removed
    - $u \in \mathcal{P}$
- Used to find interesting units

## Math Formula

For $u \in \mathcal{P}$, the change in attribute value can be written as:
$$
\begin{aligned}
\Delta(a,u) = &a(y_1, y_2, ..., y_{u-1}, y_{u}, y_{u+1}, ..., y_n)- \\
              &a(y_1, y_2, ..., y_{u-1}, y_{u+1}, ..., y_n)
\end{aligned}
$$



## Math Example

The influence on the range of the population $(1,3,5,7,9)$ when $9$ is removed is:

$$
\begin{aligned}
  \text{influence} &= \text{range}(1,3,5,7,9)-\text{range}(1,3,5,7) \\
  &= 8 - 6 \\
  &= 2
\end{aligned}
$$

## General Code

```{r}
influence <- function(attr) {
  delta = rep(0, length(pop))
  pop.attr <- attr(pop)
  for (i in 1:length(pop)) {
    delta[i] = pop.attr - attr(pop[-i])
  }
  return(delta)
}
```

## Code Example

Influence on standard deviation:
    
\small
```{r}
pop <- 1:100
pop.infl <- influence(sd)

plot(pop.infl, pch=19, xlab="Unit", ylab="Influence", 
     main="Effect on Standard Deviation")
```
---