---
title: "A3Q3"
author: "Fenglin Chen"
date: "15/11/2021"
output: word_document
---
```{r}
knitr::opts_chunk$set(fig.width=10, fig.height=5) 
```

\newpage
### Question 3
#### Description of the Context

This dataset contains class sizes for five different boroughs in New York City: Brooklyn, Manhattan, Queens, Staten Island, and Bronx. The data is collected from 2010 to 2011, and is aggregated by location, grade, and program type.

The two sub-populations of interest are the average class sizes of General Education (GEN ED) and Self-Contained Special Education (SPEC ED).

#### Compare the Sub-Populations

```{r}
boroughs <- read.csv("Borough_Summary.csv", header=TRUE)
pop <- list(pop1 = subset(boroughs, boroughs$PROGRAM.TYPE == "GEN ED"), 
            pop2 = subset(boroughs, boroughs$PROGRAM.TYPE == "SPEC ED"))
```

```{r}
qvals <- sort(pop[[1]]$AVERAGE.CLASS.SIZE)
pvals <- ppoints(length(qvals))
plot(pvals, qvals, pch = 19, col = adjustcolor("black", alpha = 0.5), 
     xlim = c(0,1), ylim = c(0, 35), 
     xlab = "Proportion p", ylab = bquote("Quantiles Q"["y"] * "(p)"), 
     main = "General Ed vs. Special Ed Class Sizes")
qvals <- sort(pop[[2]]$AVERAGE.CLASS.SIZE)
pvals <- ppoints(length(qvals))
points(pvals, qvals, pch = 19, col = adjustcolor("red", alpha = 0.5))
```

The quantile plot shows a clear separation between the sub-populations.

The functions `getAveDiffsFn()`, `getSDRatioFn()`, and `mixRandomly()` from class are used to compare the sub-populations.

```{r, echo=FALSE}
getAveDiffsFn <- function(variate) {
  function(pop) {
    mean(pop$pop1[, variate]) - mean(pop$pop2[, variate])
  }
}

getSDRatioFn <- function(variate) {
  function(pop) {
    sd(pop$pop1[, variate])/sd(pop$pop2[, variate])
  }
}

mixRandomly <- function(pop) {
  pop1 <- pop$pop1
  n_pop1 <- nrow(pop1)
  
  pop2 <- pop$pop2
  n_pop2 <- nrow(pop2)
  
  mix <- rbind(pop1, pop2)
  select4pop1 <- sample(1:(n_pop1 + n_pop2), n_pop1, replace = FALSE)
  
  new_pop1 <- mix[select4pop1, ]
  new_pop2 <- mix[-select4pop1, ]
  list(pop1 = new_pop1, pop2 = new_pop2)
}
```

```{r}
diffAveLengths <- getAveDiffsFn("AVERAGE.CLASS.SIZE")
ratioSDLengths <- getSDRatioFn("AVERAGE.CLASS.SIZE")
round(c(diffAveLengths(pop), ratioSDLengths(pop)), 3)
```

```{r, echo=FALSE}
diffLengths <- sapply(1:5000, FUN = function(...) {
  diffAveLengths(mixRandomly(pop))})
hist(diffLengths, breaks = 20, main = "Randomly Mixed Populations", 
     xlab = "Difference in Averages", col = "lightgrey", xlim=c(-3, 15))
abline(v = diffAveLengths(pop), col = "red", lwd = 2)

ratioLengths <- sapply(1:5000, FUN = function(...) {
  ratioSDLengths(mixRandomly(pop))})
hist(ratioLengths, breaks = 20, main = "Randomly Mixed Populations", xlab = "Ratio of Standard Devations", 
     col = "lightgrey", xlim=c(0.9, 1.9))
abline(v = ratioSDLengths(pop), col = "red", lwd = 2)
```

#### Description of the Comparison

The populations have a difference in class size of 14.704 and a ratio of standard deviations of 1.803. 

The graphs show that the population attributes are very far from the distribution of randomly mixed attributes. Assuming a normal distribution, the p-value for both difference in mean and ratio of standard deviations are close to 0.

#### Conclusion

Both the numerical and graphical summaries show that it is almost impossible for the two sub-populations to be picked from a random mix, as there is sufficient evidence that "H0: pop1 and pop2 are from the same distribution" can be rejected.