---
title: "Assignment 3 Question 4 Presentation"
author: "Fenglin Chen"
date: "15/11/2021"
output: beamer_presentation
---

## Table of Contents

- Summary
- Math formula
- General code
- Code example

## Summary

4 steps to finding the significance level:

1. State the null hypothesis H0
2. Formulate a discrepancy measure $D(\mathcal{P}_1, \mathcal{P}_2)$
3. Calculate the observed discrepancy $d_{obs} = D(\mathcal{P}_1, \mathcal{P}_2)$
4. Apply $D$ to randomly shuffled sub-populations

The proportion of random discrepancies greater than $d_{obs}$ is the p-value.

## Mathematical Formula

Observed discrepancy:

$$
d_{obs} = D(\mathcal{P}_1, \mathcal{P}_2)
$$

M randomly shuffled sub-populations:

$$
(\mathcal{P}_{1,1}^\star,\mathcal{P}_{2,1}^\star), (\mathcal{P}_{1,2}^\star,\mathcal{P}_{2,2}^\star), \ldots, (\mathcal{P}_{1,M}^\star,\mathcal{P}_{2,M}^\star)
$$

p-value:

$$
\text{p-value} = \frac{1}{M}\sum_{i=1}^{M}I\left(  D(\mathcal{P}_{1,i}^\star, \mathcal{P}_{2,i}^\star ) \ge d_{obs}\right)
$$

## General Code

```{r}
# Generate randomly shuffled sub-populations
mixRandomly <- function(pop) {
  pop1 <- pop$pop1
  n_pop1 <- nrow(pop1)
  
  pop2 <- pop$pop2
  n_pop2 <- nrow(pop2)
  
  mix <- rbind(pop1,pop2)
  select4pop1 <- sample(1:(n_pop1 + n_pop2), n_pop1, 
                        replace = FALSE)
  
  new_pop1 <- mix[select4pop1,]  
  new_pop2 <- mix[-select4pop1,]
  list(pop1=new_pop1, pop2=new_pop2)
}
```

## General Code

```{r}
# Define the discrepancy measure
getDiscrepancy <- function(variate, attr) {
  function(pop) {attr(pop$pop1[, variate]) - 
                 attr(pop$pop2[,variate])}
}
```

## Code Example

Gen Ed class size vs. CTT class size

```{r}
boroughs <- read.csv("Borough_Summary.csv", header=TRUE)
pop <- list(pop1 = subset(boroughs, 
                   boroughs$PROGRAM.TYPE == "GEN ED"), 
            pop2 = subset(boroughs, 
                   boroughs$PROGRAM.TYPE == "CTT"))
set.seed(341)

discFn <- getDiscrepancy("AVERAGE.CLASS.SIZE", mean)
disc.random <- sapply(1:1000, FUN = function(...){ 
  discFn(mixRandomly(pop)) 
})

disc.prop <- mean( abs(disc.random) >= discFn(pop) )
disc.prop
```

## Code Example

```{r, echo=FALSE}
# Plot the results
hist(disc.random, breaks="FD", prob=TRUE,
     main = "Randomly Mixed Populations", xlab="Difference in Class Sizes",
     col="lightgrey")
abline(v=discFn(pop), col = "red", lwd=2)
```