---
title: "Assignment 1"
author: "Fenglin Chen"
date: "21/09/2021"
output:
  word_document: default
---

### Question 1 a)

$\alpha(\mathcal{P})$ is location invariant if $\alpha(\mathcal{P}+b) = \alpha(\mathcal{P})$.

First, substitute $\sigma^4$ into the equation. $$
\alpha(\mathcal{P}) = \frac{\frac{1}{N} \sum_{u \in \mathcal{P}}\left( y_u -\overline{y} \right)^4}{ [\frac{1}{N}\sum_{u \in \mathcal{P}}(y_u-\overline y)^2]^2 }-3
$$

Then, add a constant $b$ to all elements. Since the population mean is location equivariant, it also gains $b$. $$
\alpha(\mathcal{P}+b) = \frac{\frac{1}{N} \sum_{u \in \mathcal{P}}\left( y_u + b -(\overline{y} + b) \right)^4}{ [\frac{1}{N}\sum_{u \in \mathcal{P}}(y_u + b-(\overline y + b))^2]^2 }-3
$$

The additions cancel out. $$
\alpha(\mathcal{P}+b) = \frac{\frac{1}{N} \sum_{u \in \mathcal{P}}\left( y_u + b -\overline{y} - b \right)^4}{ [\frac{1}{N}\sum_{u \in \mathcal{P}}(y_u + b-\overline y - b)^2]^2 }-3
$$ $$
\alpha(\mathcal{P}+b) = \frac{\frac{1}{N} \sum_{u \in \mathcal{P}}\left( y_u -\overline{y} \right)^4}{ [\frac{1}{N}\sum_{u \in \mathcal{P}}(y_u -\overline y)^2]^2 }-3 = \alpha(\mathcal{P})
$$ Therefore, excess kurtosis is location invariant.

\newpage
### Question 1 b)

$\alpha(\mathcal{P})$ is scale invariant if $\alpha(m*\mathcal{P}) = \alpha(\mathcal{P})$.

Similar to part a), begin by substituting $\sigma^4$ into the equation. $$
\alpha(\mathcal{P}) = \frac{\frac{1}{N} \sum_{u \in \mathcal{P}}\left( y_u -\overline{y} \right)^4}{ [\frac{1}{N}\sum_{u \in \mathcal{P}}(y_u-\overline y)^2]^2 }-3
$$ Then, multiply all elements by $m$. Since the population mean is scale equivariant, it is also multiplied by $m$. $$
\alpha(m*\mathcal{P}) = \frac{\frac{1}{N} \sum_{u \in \mathcal{P}}\left( m*y_u - m*\overline{y} \right)^4}{ [\frac{1}{N}\sum_{u \in \mathcal{P}}(m*y_u-m*\overline y)^2]^2 }-3
$$ Factor out the multiplier $m$ in both the numerator and denominator. $$
\alpha(m*\mathcal{P}) = \frac{\frac{1}{N} \sum_{u \in \mathcal{P}}(m\left(y_u - \overline{y} \right))^4}{ [\frac{1}{N}\sum_{u \in \mathcal{P}}(m(y_u-\overline y))^2]^2 }-3
$$ $$
\alpha(m*\mathcal{P}) = \frac{\frac{1}{N} \sum_{u \in \mathcal{P}}m^4\left( y_u - \overline{y} \right)^4}{ [\frac{1}{N}\sum_{u \in \mathcal{P}}m^2(y_u-\overline y)^2]^2 }-3
$$ $m$ is a constant, so it can be factored out of the sum and cancelled. $$
\alpha(m*\mathcal{P}) = \frac{m^4\frac{1}{N} \sum_{u \in \mathcal{P}}\left( y_u - \overline{y} \right)^4}{ m^4[\frac{1}{N}\sum_{u \in \mathcal{P}}(y_u-\overline y)^2]^2 }-3
$$ $$
\alpha(m*\mathcal{P}) = \frac{\frac{1}{N} \sum_{u \in \mathcal{P}}\left( y_u - \overline{y} \right)^4}{ [\frac{1}{N}\sum_{u \in \mathcal{P}}(y_u-\overline y)^2]^2 }-3 = \alpha(\mathcal{P})
$$ Therefore, excess kurtosis is scale invariant.

\newpage
### Question 1 c)

Since excess kurtosis is both location invariant (part a) and scale invariant (part b), then by definition, it is location-scale invariant.

\newpage
### Question 1 d)

$\alpha(\mathcal{P})$ is scale invariant if $\alpha(\mathcal{P}^k) = \alpha(\mathcal{P})$. Population $\mathcal{P}$ is duplicated $k-1$ times, and the new population size is $kN$.
$$
\alpha(\mathcal{P}^k) = \frac{\frac{1}{kN} \sum_{u \in \mathcal{P}^k}\left( y_u -\overline{y} \right)^4}{ [\frac{1}{kN}\sum_{u \in \mathcal{P}^k}(y_u-\overline y)^2]^2 }-3
$$
$$
\alpha(\mathcal{P}^k) = \frac{\frac{1}{kN} k\sum_{u \in \mathcal{P}}\left( y_u -\overline{y} \right)^4}{ [\frac{1}{kN}k\sum_{u \in \mathcal{P}}(y_u-\overline y)^2]^2 }-3
$$
The summing operation is applied $k$ times, but the population is $k$ times larger, so the factors cancel out.
$$
\alpha(\mathcal{P}^k) = \frac{\frac{1}{N} \sum_{u \in \mathcal{P}}\left( y_u -\overline{y} \right)^4}{ [\frac{1}{N}\sum_{u \in \mathcal{P}}(y_u-\overline y)^2]^2 }-3 = \alpha(\mathcal{P})
$$
Therefore, excess kurtosis is replication invariant.

\newpage
### Question 1 e)

In this case, the equation for $\sigma^4$ is changed.
$$
\sigma_*^4 = [\frac{1}{N-1}\sum_{u \in \mathcal{P}}(y_u-\overline y)^2]^2
$$

For part b), the equation becomes:
$$
\alpha(m*\mathcal{P}) = \frac{m^4\frac{1}{N} \sum_{u \in \mathcal{P}}\left( y_u - \overline{y} \right)^4}{ m^4[\frac{1}{N-1}\sum_{u \in \mathcal{P}}(y_u-\overline y)^2]^2 }-3
$$
$$
\alpha(m*\mathcal{P}) = \frac{\frac{1}{N} \sum_{u \in \mathcal{P}}\left( y_u - \overline{y} \right)^4}{[\frac{1}{N-1}\sum_{u \in \mathcal{P}}(y_u-\overline y)^2]^2 }-3 = \alpha(\mathcal{P})
$$
Since the multiplier $m$ does not relate to the size of the population $N$, the attribute is still scale invariant (no change). However, for part d), the equation becomes:
$$
\alpha(\mathcal{P}^k) = \frac{\frac{k}{kN} \sum_{u \in \mathcal{P}}\left( y_u -\overline{y} \right)^4}{ [\frac{k}{kN-1}\sum_{u \in \mathcal{P}}(y_u-\overline y)^2]^2 }-3
$$
$$
\alpha(\mathcal{P}^k) = \frac{\frac{1}{N} \sum_{u \in \mathcal{P}}\left( y_u -\overline{y} \right)^4}{ [\frac{1}{N-\frac{1}{k}}\sum_{u \in \mathcal{P}}(y_u-\overline y)^2]^2 }-3 \ne \alpha(\mathcal{P})
$$
Therefore, $\alpha(\mathcal{P})$ is no longer replication invariant, but rather neither invariant nor equivariant.


\newpage
### Question 1 f)

The sensitivity curve is defined as: $SC(y) = N(\alpha(\mathcal{P}^*) - \alpha(\mathcal{P}))$
$$
\alpha(\mathcal{P}^*) = \frac{\frac{1}{N} \sum_{u \in \mathcal{P}^*}\left( y_u -\overline{y} \right)^4}{ [\frac{1}{N}\sum_{u \in \mathcal{P}^*}(y_u-\overline y)^2]^2 }-3
$$
$$
\alpha(\mathcal{P}) = \frac{\frac{1}{N-1} \sum_{u \in \mathcal{P}}\left( y_u -\overline{y} \right)^4}{ [\frac{1}{N-1}\sum_{u \in \mathcal{P}}(y_u-\overline y)^2]^2 }-3
$$
Combined, the equation becomes:
$$
SC(y) = N[(\frac{\frac{1}{N} \sum_{u \in \mathcal{P}^*}\left( y_u -\overline{y} \right)^4}{ [\frac{1}{N}\sum_{u \in \mathcal{P}^*}(y_u-\overline y)^2]^2 }-3) - (\frac{\frac{1}{N-1} \sum_{u \in \mathcal{P}}\left( y_u -\overline{y} \right)^4}{ [\frac{1}{N-1}\sum_{u \in \mathcal{P}}(y_u-\overline y)^2]^2 }-3)]
$$
$$
SC(y) = N(\frac{\frac{1}{N} \sum_{u \in \mathcal{P}^*}\left( y_u -\overline{y} \right)^4}{ [\frac{1}{N}\sum_{u \in \mathcal{P}^*}(y_u-\overline y)^2]^2 } - \frac{\frac{1}{N-1} \sum_{u \in \mathcal{P}}\left( y_u -\overline{y} \right)^4}{ [\frac{1}{N-1}\sum_{u \in \mathcal{P}}(y_u-\overline y)^2]^2 })
$$



\newpage
### Question 1 g)
   
```{r}
library(e1071)
set.seed(341)
pop <- rt(1000, 10)
y <- seq(-10, 10, length.out=1001)

sc <- function(y.pop, y, attr, ...) {
  N <- length(y.pop) + 1
  sapply(y, function(y.new) {
    N * (attr(c(y.new, y.pop), ...) - attr(y.pop, ...))
  })
}

plot(y, sc(pop, y, kurtosis), type="l", lwd=2,
     main="Sensitivity Curve for Excess Kurtosis",
     ylab="sensitivity", xlab="y")
```
   
Based on this plot, kurtosis is robust to additions because its sensitivity curve is relative flat within [-5, 5]. However, it is unbounded because the two ends go to infinity, so it can be easily influenced by extreme outliers in the population.


\newpage
### Question 1 h)

1. $\gamma(\mathcal P)$ measures how close values lie in relation to each other, not in relation to some constant like 0, so it should be location invariant. A cluster of points are just as spread out at $\overline{y}=0$ as $\overline{y}=b$.
    
2. $\gamma(\mathcal P)$ should be scale equivariant. When a population of points are each multiplied by $m$, the distance/spread between any two points are also multiplied by $m$, so $\gamma(\mathcal P)$ should reflect this change linearly.

3. $\gamma(\mathcal P)$ should not be location-scale invariant nor equivariant, because location and scale have different ideal properties as seen in points 1 and 2. Choosing either would compromise location invariance or scale equivariance.


\newpage
### Question 2 a)

```{r}
filename <- paste("data/iris.csv", sep="/")
iris <- read.csv(filename, header=TRUE)
nrow(iris)
ncol(iris)
```

\newpage
### Question 2 b)

```{r}
library(pander)
iris.freq <- table(iris$Species)
pander(iris.freq, type = 'grid')
```

\newpage
### Question 2 c)

```{r}
print(paste0("Species with the largest sepal widths: ", iris$Species[iris$SepalWidth == max(iris$SepalWidth)][1]))
print(paste0("Species with the smallest sepal widths: ", iris$Species[iris$SepalWidth == min(iris$SepalWidth)][1]))
```

\newpage
### Question 2 d)

```{r}
for (s in unique(iris$Species)){
  print(paste0("Average sepal length for ", s, ": ", mean(iris$SepalLength[iris$Species == s])))
}
```

\newpage
### Question 2 e)
```{r}
iris$PetalRatio <- iris$PetalWidth/iris$PetalLength

# Iris species with largest and smallest PetalRatios
print(paste0("Species with the largest petal ratio: ", iris$Species[iris$PetalRatio == max(iris$PetalRatio)][1]))
print(paste0("Species with the smallest petal ratio: ", iris$Species[iris$PetalRatio == min(iris$PetalRatio)][1]))

# Proportion of samples with PetalRatio > 0.3
for (s in unique(iris$Species)){
  print(paste0("Proportion of samples in ", s, " with PetalRatio > 0.3: ", mean(iris$PetalRatio[iris$Species == s] > 0.3)))
}
```

\newpage
### Question 2 f)
```{r}
# Match each species name to a colour
species <- unique(iris$Species)
iris$SpeciesNum <- match(iris$Species, species)
colours <- c("black", "red", "green")

# Plot the points and add a legend
plot(iris$PetalLength, iris$SepalLength, pch=19,
     col=colours[as.numeric(iris$SpeciesNum)], 
     xlab="petal length", ylab="sepal length", 
     main="Petal Length vs. Sepal Length")
legend(1, 7.75, legend=species, col=colours, lty=1, cex=0.8)
```
   
From the plot, it is clear that Iris-setosas can be distinguished from Iris-versicolor and Iris-virginica by petal length alone, as there is a clear separation between the groups. 
    
In addition, there is a positive linear trend between petal length and sepal length in Iris-versicolor and Iris-virginica.


\newpage
### Question 2 g)

```{r}
par(mfrow=c(1,3), mar=c(2, 2, 2, 1), oma=c(4, 4, 0, 0))
for (s in species) {
  plot(jitter(PetalWidth, factor=0.5) ~ jitter(PetalLength, factor=0.5), 
       data=iris[iris$Species == s, ],
       col=adjustcolor("black", 0.3), pch=19, main=s)
  points(mean(PetalWidth) ~ mean(PetalLength), pch=19,
         data=iris[iris$Species == s, ], 
         col=adjustcolor("red", 0.5))
}
mtext('petal width', side=1, outer=TRUE, line=2)
mtext('petal length', side=2, outer=TRUE, line=2)
```
   
The plots show that there is a slight positive linear relationship between petal width and petal length, especially in the plot of Iris-versicolor. The relationship might be clearer if the values were not discrete/overlapping.

\newpage
### Question 2 h)

```{r}
# Plot the base points
par(mfrow=c(1,1))
plot(iris$SpeciesNum, iris$SepalLength, pch=19, 
     col=adjustcolor("black", 0.3), xaxt="n", 
     xlab="species", ylab="sepal length", main="Sepal Length by Species")
axis(side=1, at=c(1.0, 2.0, 3.0), labels=species)

# Calculate, plot, and connect the medians
medians <- c(median(iris$SepalLength[iris$SpeciesNum == 1]),
             median(iris$SepalLength[iris$SpeciesNum == 2]),
             median(iris$SepalLength[iris$SpeciesNum == 3]))
points(1:3, medians, col="red", pch=19)
lines(1:3, medians, col="red", pch=19)
```

\newpage
### Question 2 i)

```{r}
powerfun <- function(x, alpha) {
  if(sum(x <= 0) > 1) stop("x must be positive")
  if (alpha == 0)
    log(x)
  else if (alpha > 0) {
    x^alpha
  } else -x^alpha
}
```
```{r}
# Part i: Histogram and scatterplots of SepalLength and PetalRatio
par(mfrow=c(1,3), mar=c(2, 2, 2, 0.2))
hist(iris$SepalLength, breaks=15, main="Sepal Length")
hist(iris$PetalRatio, breaks=15, main="Petal Ratio")
plot(SepalLength ~ PetalRatio, data=iris, pch=19,
     col=adjustcolor("black", 0.5),
     main="Sepal Length vs. Petal Ratio")
```

\newpage
```{r eval=FALSE}
# Part ii: Find alpha to make SepalLength symmetric
for (i in c(-0.2, -0.1, 0, 0.1, 0.2)) {
  hist(powerfun(iris$SepalLength, i), breaks=8)
}
```

```{r}
hist(powerfun(iris$SepalLength, 0), breaks=8,
     xlab="log(Sepal Length)", ylab="Frequency", main="")
```
    
After testing, the power 0 (log) makes sepal length most symmetric.

\newpage
```{r eval=FALSE}
# Part iii: Find alpha to make PetalRatio symmetric
for (i in c(1.25, 1.5, 1.75, 2, 2.25)) {
  hist(powerfun(iris$PetalRatio, i), breaks=30)
}
```

```{r}
hist(powerfun(iris$PetalRatio, 1.75), breaks=15,
     xlab="Petal Ratio ^ 1.75", ylab="Frequency", main="")
```
    
A power of 1.75 makes the histogram of petal ratio most symmetric, though it is harder to judge than sepal length due to the second peak on the left.

\newpage
```{r}
# Part iv: Find the pair of alphas to make the scatterplot approximately linear
plot(powerfun(SepalLength, 0) ~ powerfun(PetalRatio, 1.75), data=iris, 
     pch=19, col=adjustcolor("black", 0.5),
     main="Power-Transformed Scatterplot", 
     xlab="log(Sepal Length)", ylab="Petal Ratio ^ 1.75")
```

The new scatterplot uses the best alphas for sepal length (0) and petal ratio (1.75), respectively, resulting in a more linear graph than the one in part i).

\newpage
### Question 3

```{r}
drawBoxPlot <- function(df) {
  width = 0.5
  for (i in 1:ncol(df)){
    bstats = boxplot.stats(df[,i])
    stats = bstats$stats
    
    # Plot the box and whiskers
    for ()
    
    # Plot the outliers (if they exist)
    out = bstats$out
    if (length(out) > 0){
      for (k in 1:length(out)){
        print(c(i, out[k]))
      }
    }
  }
}

# Demonstrate the function with the iris dataset
drawBoxPlot(iris[c("SepalLength", "SepalWidth", "PetalLength", "PetalWidth")])
```

```{r}
boxplot(iris[c("SepalLength", "SepalWidth", "PetalLength", "PetalWidth")])
```